Pillar 1 - Sequential Pattern Mining

### 2019-02-06 Pillar 1 - Sequential Pattern Mining.Rmd
not_taken --> fail rules

  # Spread along rounded grades
  mutate(
    grade_ceil = ceiling(Grade),
    Values     = TRUE
    ) %>%
  spread(
    key   = grade_ceil, 
    value = Values
    ) %>%

@@

  # Gather along rounded grades
  gather(
    key   = grade_ceil,
    value = X,
    `0`, `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, 
    na.rm   = TRUE, 
    convert = TRUE
    ) %>%
  select(
    -X
    ) %>%

  # Identifying item
  mutate(
    item_G = paste(item, grade_ceil, sep = "_")
    )
```
### Additional General Dataframes for courses
#### Transcript with Preceding courses per student
```{r}
d_cumulative_transcript <-  d_transactions        %>%
  select(sequenceID, item, eventID)  %>%
  arrange(sequenceID, eventID, item) %>%
  group_by(sequenceID, eventID)      %>%
  summarize_all(paste, collapse=" ") %>%
  arrange(sequenceID, eventID)       %>%
  mutate(
    course_all  = paste(unique(d_transactions$item), collapse = " "),
    course_all  = str_split(course_all, pattern = " "),

@@

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$sequenceID),
  item       = unique(d_transactions$item),
  stringsAsFactors = FALSE
  ) %>%
  full_join(d_past_student_course, by= c("sequenceID", "item")) %>%  #ATTENTION: Duplicates resits. Adds values per retake of course (eg.10177849 SCI3048)
  arrange(sequenceID) %>%
  ungroup() %>%
  drop_na(eventID) %>%
  unnest(course_past) %>%
  distinct() %>%
  group_by(item) %>%
  summarise(full_past = paste( course_past, collapse = " ")) %>% #difference, between this and above?
  mutate(full_past    = str_split(full_past, pattern = " "),
         full_past    = map(.f= unique, .x=full_past),
         total_past_courses = map(.f=length, .x=full_past))
```

@@

#
# Number of students
n_students <- d_transactions %>%
  select(sequenceID) %>%
  n_distinct

#
# Probability of taking a course
d_support_TNT <- d_transactions %>%
  distinct(
    item,
    sequenceID
    ) %>%
  count(
    item
    ) %>%
  mutate(
    rate.take = n / n_students,
    rate.not  = 1 - rate.take
    )

#
# Probability of failing, having a low grade
d_support_PF_HL <- d_transactions %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low")
    )

#
# Probability oh having a grade equal or lower than x
d_support_G <- d_transactions %>%
  group_by(
    item
    ) %>%
  summarise(
    `0`    = mean(Grade <= 0),
    `1`    = mean(Grade <= 1),
    `2`    = mean(Grade <= 2),
    `3`    = mean(Grade <= 3),
    `4`    = mean(Grade <= 4),
    `5`    = mean(Grade <= 5),
    `6`    = mean(Grade <= 6),
    `7`    = mean(Grade <= 7),
    `8`    = mean(Grade <= 8),
    `9`    = mean(Grade <= 9),
    `10`   = mean(Grade <= 10)
    ) %>%
  gather(
    key   = grade_ceil,
    value = support.grade,
    `0` : `10`,
    convert = TRUE
    )

@@

make_transaction <- function(data = d_transactions, item = item){

@@

#
# Checking transactions
# inspect(head(transactions, 10))

@@

make_sequence <- function(data = d_transactions, item = item){

  sequences <- data %>%
    filter(
      !is.na(eventID) #this is here casue we have courses that were never taken.
      ) %>% 
    group_by(
      sequenceID,
      eventID
    ) %>%
    summarise(
      list_item = list(!!enquo(item))
    ) %>%
    arrange(
      sequenceID,
      eventID
    ) %>%
    ungroup %>%
    pull(
      list_item
      ) %>%
    as("transactions")

  # indicating sequence ID and event ID for each transaction
  sequences@itemsetInfo <- select(
    data,
    sequenceID,
    eventID
    ) %>%
    as.data.frame

@@

      # include all AR possible
      parameter = list(
        supp   = 0, # min support
        smax   = 1, # max support
        conf   = 0, # min confidence
        minlen = 2, # min length of rule
        maxlen = 2  # max length of rule
        ),

@@

@@

    # rhs.support
    mutate(
      rhs_course = str_remove_all(rhs, pattern = "_fail|_pass|_low|_not taken")
      ) %>%
    left_join(
      select(data_support, item, !!enquo(type_rule)),
      by = c("rhs_course" = "item")
      ) %>%
    mutate(
      rhs.support = !!enquo(type_rule)
      )

@@

We encapsulate the four functions `my_apriori()`, `clean_AR()`, `compute_support()` and `editing_AR()` into the function `make_AR()`.

@@

```{r AR take}
AR_taken <- transactions %>%
  make_AR(
    data_support = d_support_take, 
    type_rule = rate.take
    )

rm(transactions)
```

```{r AR fail}
AR_PF <- transactions_PF %>%

  make_AR(
    data_support = d_support,
    type_rule = rate.fail
    ) %>%

  # Confidence and Lift by hand
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

@@

rm(transactions_PF)
```

```{r AR low}
AR_HL <- transactions_HL %>%

  make_AR(
    data_support = d_support,
    type_rule = rate.low
    ) %>%

  # Confidence and Lift by hand
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

@@

rm(transactions_HL)
```

```{r AR TPF}
AR_TPF <- transactions_TPF %>%
  make_AR(
    data_support = d_support, 
    type_rule = rate.fail
  ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  mutate(
    rhs.take = str_detect(rhs, "fail|pass")
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support * rhs.take),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

@@

rm(transactions_TPF)
```

```{r AR grade}
AR_G <- transactions_G %>%

  my_apriori(appearance = NULL) %>%
  clean_AR %>%

>>>>

  # Spread along rounded grades
  mutate(
    Round_grade = round(Grade, 0),
    Values = TRUE
    ) %>%
  spread(
    key   = Round_grade, 
    value = Values
    ) %>%

@@

  # Gather along rounded grades
  gather(
    key = Grade_round,
    value = X,
    `0`, `1`, `2`, `3`, `4`, `5`, `6`, `7`, `8`, `9`, `10`, 
    na.rm = TRUE, 
    convert = TRUE
    ) %>%
  select(
    -X
    )%>%

  # Identifying item
  mutate(
    item_G = paste(item, Grade_round, sep = "_")
    )
```
### Additional General Dataframes for courses
#### Transcript with Preceding courses per student
```{r}
d_cumulative_transcript <-  d_transactions        %>%
  select(sequenceID, item, eventID)  %>%
  arrange(sequenceID, eventID, item) %>%
  group_by(sequenceID, eventID)      %>%
  summarize_all(paste, collapse=" ") %>%
  arrange(sequenceID, eventID)       %>%
  mutate(
    course_all  = paste(unique(d_transactions$item), collapse = " "),
    course_all  = str_split(course_all, pattern = " "),

@@

  # Expand along students (sequenceID) and courses (itemID)
  sequenceID = unique(d_transactions$sequenceID),
  item       = unique(d_transactions$item),
  stringsAsFactors = FALSE
  ) %>%
  full_join(d_past_student_course, by= c("sequenceID", "item")) %>%  #ATTENTION: Duplicates resits. Adds values per retake of course (eg.10177849 SCI3048)
  arrange(sequenceID) %>%
  ungroup() %>%
  drop_na(eventID) %>%
  unnest(course_past) %>%
  distinct() %>%
  group_by(item) %>%
  summarise(full_past = paste( course_past, collapse = " ")) %>% #difference, between this and above?
  mutate(full_past    = str_split(full_past, pattern = " "),
         full_past    = map(.f= unique, .x=full_past),
         total_past_courses = map(.f=length, .x=full_past))
```
### Manual rules
```{r}
not_take_fail_rules <- d_cumulative_transcript     %>%
  select(sequenceID, eventID, item, course_not)    %>%
  unnest(item, .drop = F)                          %>%
  unnest(course_not)                               %>%
  left_join(
    select(d_transactions, 
           sequenceID, eventID, item, PF), 
    by= c("sequenceID", "eventID", "item")
    )                                              %>%
  select(sequenceID,eventID, course_not, item, PF) %>%
  rename(
    lhs="course_not", 
    rhs.x = "item", 
    rhs.PF = "PF")                                 %>%
  mutate(num_transactions = n())                   %>%
  group_by(lhs)                                    %>%
  mutate(
    num_lhs = n(),
    support_lhs         = num_lhs / num_transactions
    )                                              %>%
  group_by(rhs.x)                                  %>%
  mutate(
    num_rhs.x           = n(),
    support_rhs.x       = num_rhs.x / num_transactions
         )                                         %>%
  group_by(rhs.x,rhs.PF)                           %>%
  mutate(
    num_rhs.PF          = n(),
    support_rhs.PF      = num_rhs.PF / num_transactions
  )                                                %>%
  group_by(lhs, rhs.x)                             %>%
  mutate(
    num_rule_x          = n(),
    support_rule_rhs.x  = num_rule_x / num_transactions
  )                                                %>%
  group_by(lhs, rhs.x, rhs.PF)                     %>%
  mutate(
    num_rule_PF          = n(),
    support_rule_rhs.PF  = num_rule_PF / num_transactions,
    support_rhs_bounded  = num_rule_PF / num_rule_x
         )                                         %>%
  ungroup()                                        %>%
  mutate(
    confidence                   = support_rule_rhs.PF / support_lhs,
    confidence_bounded           = support_rhs_bounded / support_lhs, #not sure about this 
    lift                         = confidence / support_rhs.PF,
    lift_bounded                 = confidence_bounded /support_rhs_bounded #not sure about this
  )
```

@@

# Number of students
n_students <- d_transactions %>%
  select(sequenceID) %>%
  n_distinct()

# Probability of taken a course
d_support_take <- d_transactions %>%
  distinct(
    item,
    sequenceID
    ) %>%
  group_by(
    item
    ) %>%
  summarise(
    count     = n(),
    rate.take = count / n_students,
    rate.not  = 1 - rate.take
    )

# Probability of failing, having a low grade, having a grade less than x per course
d_support <- d_transactions %>%
  mutate(
    grade_round = round(Grade)
  ) %>%
  group_by(
    item
    ) %>%
  summarise(
    rate.fail = mean(PF == "fail"),
    rate.low  = mean(HL == "low"),
    rate.0    = mean(grade_round <= 0),
    rate.1    = mean(grade_round <= 1),
    rate.2    = mean(grade_round <= 2),
    rate.3    = mean(grade_round <= 3),
    rate.4    = mean(grade_round <= 4),
    rate.5    = mean(grade_round <= 5),
    rate.6    = mean(grade_round <= 6),
    rate.7    = mean(grade_round <= 7),
    rate.8    = mean(grade_round <= 8),
    rate.9    = mean(grade_round <= 9),
    rate.10   = mean(grade_round <= 10)
    )

@@

#
# For convenience
make_transaction <- function(data = d_transactions, item = item){

@@

#
# Checking transactions
#inspect(head(transactions_Grade, 10))

@@

#
# For convenience
make_sequence <- function(data = d_transactions, item = item){

  sequences <- data %>%
    filter(!is.na(eventID))%>% #this is here casue we have courses that were never taken.
    group_by(
      sequenceID,
      eventID
    ) %>%
    summarise(
      list_item = list(!!enquo(item))
    ) %>%
    arrange(
      sequenceID,
      eventID
    ) %>%
    ungroup %>%
    pull(
      list_item
      ) %>%
    as("transactions")

  sequences@itemsetInfo <- select(
    data,
    sequenceID,
    eventID
    ) %>%
    as.data.frame

@@

      # Chosen parameters
      parameter = list(
        supp = 0,    # min support
        smax = 1,    # max support
        conf = 0,    # min confidence
        minlen = 2,  # min length of 
        maxlen = 2  # max length of rule
        ),

@@

  type_rule_enquo <- enquo(type_rule)

@@

    # rhs.support
    mutate(
      rhs_course = str_remove_all(rhs, pattern = "_fail|_pass|_low|_not taken")
      ) %>%
    left_join(
      select(data_support, item, !!type_rule_enquo),
      by = c("rhs_course" = "item")
      ) %>%
    mutate(
      rhs.support = !!type_rule_enquo
      )

@@

We encapsulate the three functions `my_apriori()`, `clean_AR()` and `compute_support` into the function `make_AR()`.

@@

```{r AR, cache = TRUE}

#
# AR taken (confidence and lift already correct)
AR_taken <- transactions %>%
  make_AR(
    data_support = d_support_take, 
    type_rule = rate.take
    )

#
# AR fail
AR_PF <- transactions_PF %>%
  make_AR(
    data_support = d_support,
    type_rule = rate.fail
    ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

@@

#
# AR low
AR_HL <- transactions_HL %>%
  make_AR(
    data_support = d_support,
    type_rule = rate.low
    ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / rhs.support
    ) %>%
  ungroup %>%

@@

#
# AR not taken / fail / pass

AR_TPF <- transactions_TPF %>%
  make_AR(
    data_support = d_support, 
    type_rule = rate.fail
  ) %>%

  # Confidence and Lift
  separate(
    rhs,
    sep = "_",
    into = c("rhs_course", "rhs_outcome"),
    remove = FALSE
    ) %>%
  mutate(
    rhs.take = str_detect(rhs, "fail|pass")
    ) %>%
  group_by(
    lhs, rhs_course
    ) %>%
  mutate(
    lhs.rhsTake.support = sum(support * rhs.take),
    confidence          = support / lhs.rhsTake.support,
    lift                = confidence / (rhs.support + 1e-4) # some courses have a failure rate of 0%.
    ) %>%
  ungroup %>%

@@

rm(transactions, transactions_PF, transactions_HL, transactions_TPF)
```

```{r AR grade (by hand)}

#
# d_support
d_support_grade <- d_support %>%
  gather(
    key   = grade_round,
    value = support.grade,
    rate.0, rate.1, rate.2, rate.3, rate.4, rate.5, rate.6, rate.7, rate.8, rate.9, rate.10
    ) %>%
  mutate( 
    grade_round = substr(grade_round, start = 6, stop = 7),
    grade_round = as.numeric(grade_round)
    ) %>%
  select(-c(rate.fail, rate.low))

# Make AR
AR_G <- transactions_G %>%
  my_apriori(appearance = NULL) %>%
  clean_AR %>%
